Implications of uniformly distributed, empirically informed priors for phylogeographical model selection: A reply to Hickerson et al.

### 2014-07-29 manuscripts/reply/reply-si-figures.tex
More edits to reply ms.

\siFigure{../../model-choice/priors-for-pc-plot/pc-plots.pdf}{
    The prior predictive graphical checks recommended by \citet{Hickerson2013}
    for six prior models:
    (A) $M_1$ ($\divt{} \sim U(0, 0.1)$),
    (B) $M_{1A}$ ($\divt{} \sim U(0, 0.01)$),
    (C) $M_{1B}$ ($\divt{} \sim U(0, 0.001)$), 
    (D) $M_3$ ($\divt{} \sim U(0, 5)$),
    (E) $M_4$ ($\divt{} \sim U(0, 10)$), and
    (F) $M_5$ ($\divt{} \sim U(0, 20)$).
    The three models that likely exclude true values of some divergence times
    of the 22 pairs of Philippine vertebrate taxa (A--C) appear to have a
    better ``fit'' than the priors that likely cover the true divergence times
    (D--F).
    The plots project the summary statistics from 1000 random samples from each
    model onto the first two orthogonal axes of a principle component analysis,
    with the blue dot representing the observed summary statistics from the 22
    population pairs of Philippine vertebrates.
}{figPCA}

\siSidewaysFigure{../../model-choice/results/power-1/pymsbayes-results/plots/power-accuracy-6.pdf}{
    The accuracy of (A--F) unadjusted and (G--L) GLM-adjusted estimates of
    dispersion index of divergence times (\vmratio{}) when the empirically
    informed model-averaging approach of \citet{Hickerson2013} is applied to
    simlated datasets in which divergence times of 22 pairs of populations are
    randomly drawn from the uniform distributions $\divt{} \sim U(0,
    \divt{max})$ indicated at the top of each column of plots (divergence-time
    distributions are given in units of millions of generations ago (MGA)
    assuming a per-site rate of 1\e{-8} mutations per generation).
}{figPowerAccuracy}

@@

\siSidewaysFigure{../../model-choice/results/power-1/pymsbayes-results/plots/power-exclusion-6.pdf}{
    Histograms of the (A--F) number of true divergence-time parameters excluded
    from the preferred model and the (G--L) posterior probability of excluding
    at least one divergence-time parameter when the empirically informed
    model-averaging approach of \citet{Hickerson2013} is applied to simlated
    datasets in which divergence times of 22 pairs of populations are randomly
    drawn from the uniform distributions $\divt{} \sim U(0, \divt{max})$
    indicated at the top of each column of plots (divergence-time distributions
    are given in units of millions of generations ago (MGA) assuming a per-site
    rate of 1\e{-8} mutations per generation).
}{figPowerExclusion6}

>>>>

\siFigure{../../model-choice/priors-for-pc-plot/pc-plots.pdf}{
    The prior predictive graphical checks recommended by \citet{Hickerson2013}
    for six prior models:
    (A) $M_1$ ($\divt{} \sim U(0, 0.1)$),
    (B) $M_{1A}$ ($\divt{} \sim U(0, 0.01)$),
    (C) $M_{1B}$ ($\divt{} \sim U(0, 0.001)$), 
    (D) $M_3$ ($\divt{} \sim U(0, 5)$),
    (E) $M_4$ ($\divt{} \sim U(0, 10)$), and
    (F) $M_5$ ($\divt{} \sim U(0, 20)$).
    The three models that likely exclude true values of some divergence times
    of the 22 pairs of Philippine taxa (A--C) appear to have a ``better fit''
    than the valid priors that likely cover the true divergence times
    (D--F).
    The plots project the summary statistics from 1000 random samples from each
    model onto the first two orthogonal axes of a principle component analysis,
    with the blue dot representing the observed summary statistics from the 22
    population pairs of Philippine vertebrates.
}{figPCA}

\siSidewaysFigure{../../model-choice/results/power-1/pymsbayes-results/plots/power-accuracy-6.pdf}{
    The accuracy of (A--F) unadjusted and (G--L) GLM-adjusted estimates of
    the dispersion index of divergence times (\vmratio{}) when the empirically
    informed model-averaging approach of \citet{Hickerson2013} is applied to
    simlated datasets in which divergence times of 22 pairs of populations are
    randomly drawn from the uniform distributions $\divt{} \sim U(0,
    \divt{max})$ indicated at the top of each column of plots (divergence-time
    distributions are given in units of millions of generations ago (MGA)
    assuming a per-site rate of 1\e{-8} mutations per generation).
}{figPowerAccuracy}

@@

\siSidewaysFigure{../../model-choice/results/power-1/pymsbayes-results/plots/power-exclusion-6.pdf}{
    Histograms of the (A--F) number of true divergence times excluded
    from the preferred model and the (G--L) posterior probability of excluding
    at least one true divergence time when the empirically informed
    model-averaging approach of \citet{Hickerson2013} is applied to simlated
    datasets in which divergence times of 22 pairs of populations are randomly
    drawn from the uniform distributions $\divt{} \sim U(0, \divt{max})$
    indicated at the top of each column of plots (divergence-time distributions
    are given in units of millions of generations ago (MGA) assuming a per-site
    rate of 1\e{-8} mutations per generation).
}{figPowerExclusion6}

### 2014-07-29 manuscripts/reply/reply-si-main.tex
More edits to reply ms.

The fact that their posterior samples are in different units makes the results
of \citet{Hickerson2013} difficult to interpret, and renders their
regression-adjusted results invalid.
A fundamental assumption of regression is that all of the values of the
response variable are in the same units.
Thus, the results in sections ``Using ABC Model Comparison to Weight
Alternative Priors for the Philippine Vertebrate Data'' and ``Improved Sampling
Efficiency by Prior Weighting Supports Asynchronous and Recent Divergence for
the Philippines Vertebrate Data'' and presented in Figure 2 of
\citet{Hickerson2013} should be disregarded.
The error is easily illustrated by re-plotting their results with the different
time units indicated (Figure~S\ref{figJointPosteriorHickerson}).

@@

\subsection{Saturation of summary statistics}
\citet{Hickerson2013} claim the priors used by \citet{Oaks2012} ``cause much of
the explored parameter space to be beyond the threshold of saturation in most
mtDNA genes.'' To explore this possibility, we simulated datasets under prior
settings that match two of the three priors used by \citet{Oaks2012}:
$\meanDescendantTheta{} \sim U(0.0005, 0.04)$ and $\ancestralTheta{} \sim
U(0.0005, 0.02)$.
Under this prior, we randomly sample divergence-time parameters from a uniform
distribution of $U(0, 20)$ coalescent units, simulate datasets, and plot the
\divt{} values against the summary statistics calculated from the resulting
datasets (Figure~\ref{figSaturationPlot}).
Clearly, the priors used by \citet{Oaks2012} with upper limits on \divt{} of five
and 10 coalescent units suffered little to no effect from saturation.
Even at divergence times of 20 coalescent units, there is still signal in the
summary statistics used by \msb (Figure~\ref{figSaturationPlot}).
Thus, the assertion of \citet{Hickerson2013} does not apply to at least
two of the priors used by \citet{Oaks2012} and, as a result, does not
explain the bias they found.

@@

\subsection{Differing utilities of \numt{} and \vmratio{} in \texttt{msBayes}}
The primary component of the \msb model is the vector of divergence
times for each of the taxon pairs,
$\divtvector = \{\divt{1}, \ldots, \divt{Y}\}$
\citep{Oaks2012}.
\citet{Hickerson2013} argue that the dispersion index of this vector,
\vmratio{}, is a better model-choice estimator than the number of 
divergence-time parameters within the vector,
\numt{}.
They present a plot of \numt{} against \vmratio{} (Fig.~S1 of
\citet{Hickerson2013}), which is essentially a plot of sample size versus
variance.
This plot shows, not surprisingly, that \vmratio{} has very little information
about the number of divergences among taxa.
Nonetheless, \citet{Hickerson2013} conclude \vmratio{} is more informative and
biogeographically relevant than \numt{}.
% We struggle to follow this logic.
However, all of the information about the temporal distribution of divergences
is contained within the divergence-time vector that \vmratio{} is summarizing.
Clearly, the number of divergence-time parameters within the vector and their
values is more informative than the variance (i.e., the dispersion index is not
a sufficient statistic for \divtvector).
\citet{Hickerson2013} also argue that \msb can estimate \vmratio{} much better
than \numt{}.
However, \citet{Oaks2012} demonstrate that even when all assumptions of the
model are met, \vmratio{} is a poor model-choice estimator (see plots B, D \& F
of Figure 4 in \citet{Oaks2012}), whereas \numt{} performs better.

Importantly, \vmratio{} is limited to estimating the probability of only a
single model (the one-divergence model), and thus its utility for model-choice
is very limited.
I.e., it can only be informative about the probability of whether there is one
divergence shared among the taxa ($\vmratio{} = 0.0$) or there is greater than
one divergence ($\vmratio{} > 0.0$).
As a result, not only is its model-choice utility limited, but it is also
very difficult to estimate.
\vmratio{} can range from zero to infinity, and the point density that it is
at its lower limit of zero will always be zero.
Thus, an arbitrary threshold (0.01 is used throughout the \msb literature) must
be chosen to make the probability of ``simultaneous'' divergence estimable.
Even with this arbitrary threshold, it is still not surprising to see that it
is numerically difficult to obtain reliable estimates of the probability that
\vmratio{} is ``near'' its lower limit of zero.
It is easier, less subjective, and more interpretable to estimate the
probability of the discrete parameter of the model, \numt{}, is at its lower
limit of one.
Thus, it is not surprising that \citet{Oaks2012} find that \numt{} is a better
estimator of model probability than \vmratio{}.

>>>>

A fundamental assumption of the \msb model and post hoc regression adjustment
is that all possible values of the parameter of interest (divergence times) are
in the same units.
Thus, the results in sections ``Using ABC Model Comparison to Weight
Alternative Priors for the Philippine Vertebrate Data'' and ``Improved Sampling
Efficiency by Prior Weighting Supports Asynchronous and Recent Divergence for
the Philippines Vertebrate Data'' and presented in Figure 2 of
\citet{Hickerson2013} are invalid and should be disregarded.
The error is easily illustrated by re-plotting their results with the different
time units indicated (Figure~S\ref{figJointPosteriorHickerson}).

@@

\subsection{Saturation of summary statistics}
\citet{Hickerson2013} claim the priors used by \citet{Oaks2012} ``cause much of
the explored parameter space to be beyond the threshold of saturation in most
mtDNA genes.'' To explore this possibility, we simulated datasets under prior
settings that match two of the three priors used by \citet{Oaks2012}:
$\meanDescendantTheta{} \sim U(0.0005, 0.04)$ and $\ancestralTheta{} \sim
U(0.0005, 0.02)$.
Under this prior, we randomly sample divergence-time parameters from a uniform
distribution of $U(0, 20)$ coalescent units, simulate datasets, and plot the
\divt{} values against the summary statistics calculated from the resulting
datasets (Figure~\ref{figSaturationPlot}).
Clearly, the priors used by \citet{Oaks2012} with upper limits on \divt{} of five
and 10 coalescent units suffered little to no effect from saturation.
Even at divergence times of 20 coalescent units, there is still signal in the
summary statistics used by \msb (Figure~\ref{figSaturationPlot}).
Thus, the assertion of \citet{Hickerson2013} that the priors used by
\citet{Oaks2012} sample parameter space in which the mtDNA alignments are
saturated by substitutions is incorrect and, as a result, does not explain the
bias they found.

@@

\subsection{Differing utilities of \numt{} and \vmratio{} in \texttt{msBayes}}
The primary component of the \msb model is the vector of divergence
times for each of the taxon pairs,
$\divtvector = \{\divt{1}, \ldots, \divt{Y}\}$
\citep{Oaks2012}.
\citet{Hickerson2013} argue that the dispersion index of this vector,
\vmratio{}, is a better model-choice estimator than the number of 
divergence-time parameters within the vector,
\numt{}.
They present a plot of \numt{} against \vmratio{} (Fig.~S1 of
\citet{Hickerson2013}), which is essentially a plot of sample size versus
variance.
This plot shows that \vmratio{} has very little information
about the number of divergences among taxa.
Nonetheless, \citet{Hickerson2013} conclude \vmratio{} is more informative and
biogeographically relevant than \numt{}.
% We struggle to follow this logic.
However, the number of divergence-time parameters within the vector and their
values contains all of the information about the temporal distribution of
divergences, and is much more informative than the variance (i.e., the
dispersion index is not a sufficient statistic for \divtvector).
\citet{Hickerson2013} also argue that \msb can estimate \vmratio{} much better
than \numt{}.
However, \citet{Oaks2012} demonstrate that even when all assumptions of the
model are met, \vmratio{} is a poor model-choice estimator (see plots B, D \& F
of Figure 4 in \citet{Oaks2012}), whereas \numt{} performs better.

Importantly, \vmratio{} is limited to estimating the probability of only a
single model (the one-divergence model), and thus its utility for model-choice
is very limited.
I.e., it can only be informative about the probability of whether there is one
divergence shared among the taxa ($\vmratio{} = 0.0$) or there is greater than
one divergence ($\vmratio{} > 0.0$).
As a result, not only is its model-choice utility limited, but it is also
very difficult to estimate.
\vmratio{} can range from zero to infinity, and the point density that it is
at its lower limit of zero will always be zero.
Thus, an arbitrary threshold (0.01 is used throughout the \msb literature) must
be chosen to make the probability of ``simultaneous'' divergence estimable.
Even with this arbitrary threshold, it is still not surprising to see that it
is numerically difficult to obtain reliable estimates of the probability that
\vmratio{} is ``near'' its lower limit of zero.
It is easier, less subjective, and more interpretable to estimate the
probability of the model with one divergence-time parameter (i.e., $\numt{} =
1$).
Thus, it is not surprising that \citet{Oaks2012} find that \numt{} is a better
estimator of model probability than \vmratio{}.
